babette: BEAUti 2, BEAST2 and Tracer for R

### 2018-02-09 article.tex
Better

  \textbf{1. }
    In the field of phylogenetics, 
     BEAST2 is one of the most widely used software tools. 
     It comes with the graphical programs BEAUti 2 to
     facilitate the creation of input files to BEAST2, and
     with DensiTree and Tracer 
     to interpret BEAST 2's created output files. 
     However, when many input and output files are required, 
     a workflow of GUI programs is cumbersome. 
     Moreover, many other phygenetics tools are available 
     in R, which requires switching from one platform 
     to the other. \\
  \textbf{2. }
    Here, we present a free, libre and open-source package, \verb;babette;, 
    'BEAUti 2, BEAST2 and Tracer for R', for the R programming language. 
    \verb;babette; creates BEAST2 input files, runs BEAST2 and parses its results, 
    from an R function call. \\
  \textbf{3. }
    We describe \verb;babette;'s usage, the novel functionality it provides
    compared to BEAUti 2, and give some examples. \\
  \textbf{4. }
    As \verb;babette; is designed to be of high quality and extendable, 
    we conclude by describing the further development of the package \\
\end{abstract}

@@

BEAST2 is also associated with Tracer \cite{tracer} and 
DensiTree 2 \cite{DensiTree}. Both are desktop applications 
to analyze the output of BEAST2, each with a user-friendly graphical user interface.

@@

Here, to provide such functionality we present \verb;babette;, 
’BEAUti 2, BEAST2 and Tracer for R’, 
which creates BEAST2 configuration files,
runs BEAST2 and analyzes its results 
from an R function call and hence
will save time and tedious mouse clicking and 
reduces the chances of errors in such repetitive actions.
The interface of \verb;babette; to create BEAST2 configureation
files closely mimics BEAUti. This
familiarity helps both beginner and experienced BEAST2 users 
to create configuration files using \verb;babette;.
\verb;babette; enables the creation of a single-script 
pipeline from sequence alignments to tree analysis in R.

@@

\verb;babette; has 20 exported function to parse the BEAST2 output
file and analyse the created posterior. \verb;babette; gives the
same ESSes and summary statistics as Tracer. The data is formatted
as such, that it can easily be visualized using ggplot2 (for a trace,
similar to Tracer)
or phangorn (for the phylogenies in a posterior, simular to DensiTree).

@@

Our initial motivation to create \verb;babette; 
is that we wanted to fix the crown age of a phylogeny.
BEAUti assumes that a phylogeny has a crown age that needs to be jointly-estimated
with the phylogeny and other parameters. It does not allow for fixing
the crown age. Without \verb;babette;, one needs to manually edit the BEAST2 
XML configuration file, which is tedious and prone to errors. 
Fixing the crown ages allows for theoretical experiments 
with one less source of variation.

@@

\verb;babette; has functionality of both Tracer and
DensiTree. To calculate the effective sample sizes in \verb;babette;:

\begin{lstlisting}[language=R, caption=Calculate ESSes, label=lst:calc_esses, floatplacement=H]
traces <- remove_burn_ins(traces = out$estimates, burn_in_fraction = 0.2)
esses <- calc_esses(traces, sample_interval = 1000)
\end{lstlisting}

>>>>

  \textbf{1. }
    In the field of phylogenetics, 
     BEAST2 is one of the most widely used software tools. 
     It comes with the graphical programs BEAUti 2 to
     facilitate the creation of input files to BEAST2, and
     with DensiTree and Tracer, 
     to interpret BEAST2's output. 
     However, when many input files are required, 
     a workflow of GUI programs is cumbersome. 
     Moreover, many other phygenetics tools are available 
     in R, which requires switching from one platform 
     to the other. \\
  \textbf{2. }
    Here, we present a free, libre and open-source package, \verb;babette;, 
    'BEAUti 2, BEAST2 and Tracer for R', for the R programming language. 
    \verb;babette; creates BEAST2 input files, runs BEAST2 and parses its results, 
    from an R function call. \\
  \textbf{3. }
    We describe \verb;babette;'s usage, the novel functionality it provides
    compared to the tools it replaces, and give some examples. \\
  \textbf{4. }
    As \verb;babette; is designed to be of high quality and extendable, 
    we conclude by describing the further development of the package \\
\end{abstract}

@@

BEAST2 is also associated with Tracer \cite{tracer} and 
DensiTree \cite{DensiTree}. Both are desktop applications 
to analyze the output of BEAST2, each with a user-friendly graphical user interface.

@@

Here, to provide such functionality we present \verb;babette;, 
’BEAUti 2, BEAST2 and Tracer for R’, 
which creates BEAST2 configuration files,
runs BEAST2, and analyzes its results,
all from an R function call. This
will save time, tedious mouse clicking and 
reduces the chances of errors in such repetitive actions.
The interface of \verb;babette; mimics the tools it
is inspired on. This
familiarity helps both beginner and experienced BEAST2 users 
to make the step from those tools to \verb;babette;.
\verb;babette; enables the creation of a single-script 
pipeline from sequence alignments to posterior analysis in R.

@@

\verb;babette; has 20 exported function to parse the BEAST2 output
file and analyse the created posterior. \verb;babette; gives the
same ESSes and summary statistics as Tracer. The data is formatted
as such, that it can easily be visualized using \verb;ggplot2; (for a trace,
similar to Tracer)
or \verb;phangorn; (for the phylogenies in a posterior, simular to DensiTree).

@@

Our initial motivation to create \verb;babette; 
is that we wanted to fix the crown age of a phylogeny.
BEAUti assumes that a phylogeny has a crown age that needs to be jointly-estimated
with the phylogeny and other parameters. It does not allow for fixing
the crown age. Without \verb;babette;, one needs to manually edit the BEAST2 
configuration file, which is tedious and prone to errors. 
Fixing the crown ages is especially useful for theoretical experiments,
as this allows for one less source of variation.

@@

\verb;babette; has functionality of both Tracer and
DensiTree. Tracer, among others, shows the effective sample
sizes of each posterior's variables. As an MCMC run needs to
converge first, Tracer discards the first 10\% of all parameter
estimates. This is called the burn-in. To calculate the effective
sample sizes in \verb;babette;, with the same burn-in of 10\%:

\begin{lstlisting}[language=R, caption=Calculate ESSes, label=lst:calc_esses, floatplacement=H]
traces <- remove_burn_ins(traces = out$estimates, burn_in_fraction = 0.1)
esses <- calc_esses(traces, sample_interval = 1000)
\end{lstlisting}
